\bf Estimating the number of clusters using cross-validation

### 2016-06-16 cvclust.tex
revise two-cluster theory

\newcommand{\OhP}{O_p}

@@

\begin{theorem}\label{thm:self-consistency}

Let $\{ (X_i, Y_i) \}_{i=1}^{n+m}$ be the data from a single fold of Gabriel
cross-validation.  For any $k$, let $CV(k)$ be the cross-validation error for
this fold, computed as described in Section~\ref{sec:gabriel-cv-algorithm}.
We will assume that there are $K$ true centers $\mu(1), \dotsc,\mu(K)$, with
the $g$th cluster center partitioned as $\mu(g) = \bigl(\muX(g),
\muY(g)\bigr)$ for $g = 1, \dotsc, K$.  Suppose that
\begin{enumerate}[label=(\roman*)]
  \item \label{asn:self-consistency-noiseless}
    Each observation $i$ has a true cluster $G_i \in \{ 1, \dotsc, K \}$.
    There is no noise, so that $X_i = \muX({G_i})$ and $Y_i = \muY(G_i)$ for
    $i = 1, \dotsc, n+m$.
  \item \label{asn:self-consistency-distinct-mux}
    The vectors $\muX(1), \dotsc,\muX(K)$ are all distinct.
  \item \label{asn:self-consistency-distinct-muy}
    The vectors $\muY(1), \dotsc,\muY(K)$ are all distinct.
  \item \label{asn:self-consistency-train}
    The training set contains at least one member of each cluster: for all $g$
    in the range $1, \dotsc, K$, there exists at least one $i$ in the range
    $1, \dotsc, n$ such that $G_i = g$.
  \item \label{asn:self-consistency-test}
    The test set contains at least one member of each cluster: for all $g$ in
    the range $1, \dotsc, K$, there exists at least one $i$ in the range $n+1,
    \dotsc, n+m$ such that $G_i = g$.
\end{enumerate}
Then $CV(k) < CV(K)$ for $k < K$, and $CV(k) = CV(K)$ for $k > K$, so that
Gabriel CV correctly chooses $k = K$.
\end{theorem}

The result statement of Theorem~\ref{thm:self-consistency} follows from 
Lemmas~\ref{lem:self-consistency1} and~\ref{lem:self-consistency2}, which we
will now state and prove.

\begin{lemma}\label{lem:self-consistency1}
Suppose that the assumptions of Theorem~\ref{thm:self-consistency} are in
force.  If $k < K$, then $\CV(k) > 0$.
\end{lemma}
\begin{proof}
By definition,
\[
  \CV(k)
    =
      \sum_{i=n+1}^{n+m}
        \| Y_i - \bmuY (\hGX_i) \|^2,
\]
where $\bmuY(g)$ is the center of cluster $g$ returned from applying $k$-means
to $Y_1, \dotsc, Y_n$.  Assumptions~\ref{asn:self-consistency-noiseless}
and~\ref{asn:self-consistency-test}, imply that as $i$ ranges over the test
set $n+1, \dotsc, n+m$, the response $Y_i$ ranges over all distinct values in
$\{ \muY(1), \dotsc, \muY(K) \}$.
Assumption~\ref{asn:self-consistency-distinct-muy} implies that there are
exactly $K$ such distinct values.  However, there are only $k$ distinct values
of $\bmuY(g)$.  Thus, at least one summand
\(
  \| Y_i - \bmuY(\hGX_i) \|^2
\)
is nonzero.  Therefore,
\(
  \CV(k) > 0.
\)
\end{proof}

\begin{lemma}\label{lem:self-consistency2}
Suppose that the assumptions of Theorem~\ref{thm:self-consistency} are in
force.  If $k \geq K$, then $\CV(k) = 0$.
\end{lemma}
\begin{proof}
From assumptions~\ref{asn:self-consistency-noiseless},
\ref{asn:self-consistency-distinct-muy},
and~\ref{asn:self-consistency-train}, we know the cluster centers
gotten from applying $k$-means to $Y_1, \dotsc, Y_n$ must include
$\muY(1), \dotsc, \muY(K)$.  Without loss of generality, suppose that
$\bmuY(g) = \muY(g)$ for $g = 1, \dotsc, K$.  This implies that
$\hGY_i = G_i$ for $i = 1, \dotsc, n$.  Thus, employing
assumption~\ref{asn:self-consistency-noiseless} again, we get that
$\bmuX(g) = \muX(g)$ for $g = 1, \dotsc, K$.

@@

Theorem~\ref{thm:self-consistency} tells us that the Gabriel cross-validation
method recovers the true number of clusters when the noise is negligible.
While this result gives us some assurance that the procedure is well-behaved,
we can bolster our confidence and gain insight into its workings by analyzing
its behavior in the presence of noise.  We first study the case of a single
cluster in two dimensions with correlated Gaussian noise.

\begin{theorem}
Suppose that $\{ (X_i, Y_i) \}_{i=1}^{n + m}$ is data from a single fold
of Gabriel cross-validation, where each $(X,Y)$ pair in $\R^2$ is an
independent draw from a mean-zero multivariate normal distribution with unit
marginal variances and correlation $\rho$.  In this case, the data are drawn
from a single cluster; the true number of clusters is~$1$.  If $|\rho| < 1/2$,
and $k > 1$, then $\CV(1) < \CV(k)$ with probability tending to one as $m$
and $n$ increase.
\end{theorem}

@@

Intuitively, when the correlation is high, the response feature, $Y$, looks
similar to the predictor feature, $X$. Prediction error on $X$ always
decreases with larger $k$. Thus, when the correlation is high, the prediction
error for $Y$ will also decrease with larger $k$. This explains why
cross-validation breaks down in the presense of strong correlation.

\subsection{Single Cluster in higher dimensions}

Let $\Sigma_{YY} = U \Lambda U^T$  be the eigendecomposition of $\Sigma_{YY}$, with leading eigenvalue $\lambda_1$ and corresponding eigenvector $u_1$. Then the centroid of $k$-means applying on $(y_1,..,y_n)$ is on the first PC 
of $Y$,\[	E(u^T_1 Y|u^T_1 Y>0) = \bar{\mu}^Y_1 =\sqrt{2 \lambda_1/\pi}u_1\] and 
\[	E(u^T_1 Y|u^T_1 Y<0) = \bar{\mu}^Y_2 =-\sqrt{2 \lambda_1/\pi}u_1\]
 where $u^T_1 Y \sim \mathcal{N}(0,\lambda_1)$.

To compute $\bar{\mu}^X_1 = E(X|u^T_1 Y>0)$, we need to know the conditional distribution $X|u^T_1 Y$. Since $(X,Y)$ has multivariate normal distribution, $(X,u^T_1 Y)$ also has a multivariate normal distribution with mean $\mathbf{0}$ and covariance matrix
$$\Sigma_{X,u^T_1 Y}=\begin{bmatrix} \Sigma_{XX} & \Sigma_{XY} u_1 \\  u^T_1 \Sigma_{YX} & \lambda_1 \end{bmatrix}$$
The conditional distribution $X|u^T_1 Y$ is hence normal with mean
 $$\mu_{X|u^T_1 Y} = \Sigma_{XY} u_1 \lambda^{-1}_1 u^T_1 Y $$
Therefore, 
\begin{align}
\bar{\mu}^X_1 &= E(X \mid u^T_1 Y>0) \nonumber \\ \nonumber
 			  &= E\left(E[X \mid u^T_1 Y] \mid u^T_1 Y>0\right) \\ \nonumber
 			  &=  E\left(\Sigma_{XY} u_1 \lambda^{-1}_1 u^T_1 Y \mid u^T_1 Y>0\right)\\ \nonumber
 			  &= \lambda^{-1}_1 \Sigma_{XY}u_1 E(u^T_1 Y \mid u^T_1 Y>0) \\ \nonumber
 			  &= \lambda^{-1}_1 \Sigma_{XY}u_1 \sqrt{2 \lambda_1/\pi} \\ \nonumber
 			  &= \sqrt{2 / \lambda_1 \pi} \Sigma_{XY}u_1
\end{align}
Similar calculation yields $\bar{\mu}^X_2 = -\sqrt{2 / \lambda_1 \pi} \Sigma_{XY}u_1$.
The decision rule to classify any observed value of $X$ to $\bar{\mu}^X_1$ is therefore
\[	(\bar{\mu}^X_1)^T X >0	\hspace{0.2in}\text{or} \hspace{0.2in} u^T_1\Sigma_{YX}X>0\] 
Since $u^T_1\Sigma_{YX}X$ is a linear combination of $X$, it also has normal distribution 
\[	\mathcal{N} \left( 0, u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1\right)	\]
And $(Y,u^T_1\Sigma_{YX}X)$ also have multivariate normal distribution with mean $\mathbf{0}$ 
and covariance matrix
\[
\begin{bmatrix}
\Sigma_{YY} & \Sigma_{YX}\Sigma_{XY}u_1  \\
u^T_1\Sigma_{YX}\Sigma_{XY} &  u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1
\end{bmatrix}
\]
The conditional distribution of $Y|u^T_1\Sigma_{YX}X$ is also multivariate normal with mean 
\[	
\mu_{Y|u^T_1\Sigma_{YX}X } = \Sigma_{YX}\Sigma_{XY}u_1 (u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1)^{-1}u^T_1\Sigma_{YX}X	
\]
The $Y$ center for $u^T_1\Sigma_{YX}X>0$ is
\begin{align}
\hat{\mu}^Y_1 &= E(Y|u^T_1\Sigma_{YX}X>0) \nonumber \\ \nonumber
 		      & =  \Sigma_{YX}\Sigma_{XY}u_1 (u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1)^{-1} E(u^T_1\Sigma_{YX}X \mid u^T_1\Sigma_{YX}X>0) \\ \nonumber
\end{align}
Note that $u^T_1\Sigma_{YX}X$ has normal distribution $\mathcal{N} \left( 0, u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1\right)$, so
\[
E(u^T_1\Sigma_{YX}X \mid u^T_1\Sigma_{YX}X>0) = \sqrt{2/\pi}\cdot\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1}
\]
Therefore, we have the $Y$ center for $u^T_1\Sigma_{YX}X>0$ be
\begin{align*}
\hat{\mu}^Y_1 &= \sqrt{2/\pi}\cdot\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1} \hspace{0.1in} \Sigma_{YX}\Sigma_{XY}u_1 (u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1)^{-1} \\
&=\frac{\sqrt{2/\pi}}{\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1}} \Sigma_{YX}\Sigma_{XY}u_1
\end{align*}

Recall that $\bar{\mu}^Y_1 =\sqrt{2 \lambda_1/\pi}u_1$, to judge if $CV(2) > CV(1)$, one only need to compare the distance between  $\hat{\mu}^Y_1$  and  $\bar{\mu}^Y_1$ with distance between  $\hat{\mu}^Y_1$ and grand mean $0$. By variance and bias decomposition of prediction MSE, when variance is the same, only bias influence the MSE.

After some linear algebra manipulation, we get
$||\hat{\mu}^Y_1 - \bar{\mu}^Y_1||^2 > ||\hat{\mu}^Y_1||^2$ or $CV(2) > CV(1)$ iff
\[
 \frac{\sqrt{\lambda_1}}{2} > \frac{u^T_1\Sigma_{YX}\Sigma_{XY}u_1}{\sqrt{u^T_1\Sigma_{YX} \Sigma_{XX} \Sigma_{XY} u_1}} 
\]
\end{proof}

@@

We will now analyze a simple two-cluster setting, and derive conditions for
Gabriel cross-validation to correctly prefer $k=2$ clusters to $k=1$.

\begin{proposition}
Suppose that $\{(X_i,Y_i)\}_{i=1}^{n+m}$ is data from a single fold of Gabriel
cross-validation, where each $(X,Y)$ pair in $\R^2$ is an independent draw
from an equiprobable mixture of two multivariate normal distributions with
identity covariance. Suppose that the first mixture component has mean $(\muX,
\muY)$, and the second has mean $(-\muX, -\muY)$, where $\muX > 0$ and
$\muY > 0$.  If $1 + 2\Phi(\mu^Y)+ \frac{2\varphi(\mu^Y)}{\mu^Y} < 4\Phi(\mu^X)$, then $\CV(2) < \CV(1)$ with
probability tending to one as $m$ and $n$ increase.
\end{proposition}

\begin{proof}
There are two clusters $G_1$ and $G_2$, where observations from $G_1$ are distributed as
\[	\mathcal{N}\left( \begin{pmatrix} 
    \mu^X \\
    \mu^Y \\
  \end{pmatrix}, \mathbf{I} \right)	\]
and observations from $G_2$ are distributed as
\[	\mathcal{N}\left( \begin{pmatrix} 
    -\mu^X \\
    -\mu^Y \\
  \end{pmatrix}, \mathbf{I} \right)	\]
where $\mu^X_1 > 0$ and $\mu^Y_1 > 0$. Let $G_i$ be the true cluster where observation $i$ is generated from, by assumption
\[	P(G_i=G_1) = P(G_i=G_2) = 1/2	\]
After applying $k$-means on $\{ Y_i\}_{i=1}^{n}$ with $k=2$, if $n$ is large enough, we have the estimated centroids $\bar{\mu}^Y_1$ and $\bar{\mu}^Y_2$ be close to $E(Y \mid Y>0)$ and $\E(Y \mid Y < 0)$,
with errors will be of size $\OhP(n^{-1/2})$. Here
\begin{align}
E(Y \mid Y>0) &= E(Y_1 \mid Y_1 > 0) \cdot P(Y_1>0) + E(Y_2 \mid Y_2 > 0) \cdot P(Y_2>0) \nonumber \\ 
   &= 2\varphi(\mu^Y)+2\mu^Y\Phi(\mu^Y)-\mu^Y  \\  \nonumber 
\end{align}
where $Y_1 \sim N(\mu^Y, 1)$ and $Y_2 \sim N(-\mu^Y, 1)$, and we used Lemma~\ref{lem:truncated-normal-moments}
(Appendix~\ref{app:technical-lemmas}). Similarly, we have
\begin{align}
E(Y \mid Y<0) &= E(Y_1 \mid Y_1 < 0) \cdot P(Y_1<0) + E(Y_2 \mid Y_2 < 0) \cdot P(Y_2<0) \nonumber \\ 
   &= -2\varphi(\mu^Y)-2\mu^Y\Phi(\mu^Y)+\mu^Y  \\  \nonumber 
\end{align}
 where $\varphi()$ and $\Phi()$ are the standard normal probability and cumulative distribution function respectively.

Same as in single cluster case, the classification rule learned from
$\{(X_i, \hGY_i)\}_{i=1}^{n}$ variables will be determined according to
whether $X > 0$, with the decision boundary be at $0 + \OhP(n^{-1/2})$.
By symmetry, the CV error for points from $G_1$ is same as the points from $G_2$. Because $P(G=G_1) = P(G=G_2) =1/2$, the CV error can be calculated solely from $G_2$, that is

\begin{align*}
  \CV(2)
  &=
    \frac{1}{m}
    \sum_{i=n+1}^{n+m}
      \| (Y_i - \bmuY_1) 1\{\hGX_i = 1\} \|^2
      +
      \| (Y_i - \bmuY_2) 1\{\hGX_i = 2\} \|^2, \hspace{0.2in}Y\sim N(-\mu^Y,1)
\\
  &=
    \E[(Y - a)^2 1\{ X > 0\}] + \E[(Y + a)^2 1\{X < 0\}]
      + \OhP(m^{-1/2}) + \OhP(n^{-1/2})\\
   \end{align*}
   With 
   \begin{align}
 \E[(Y - a)^2 1\{ X > 0\}] + \E[(Y + a)^2 1\{X < 0\}]  =& P(\hGX_i = 1)\cdot E[(Y - a)^2] + P(\hGX_i = 2)\cdot E[(Y + a)^2] \nonumber \\  \nonumber
 =& [1-\Phi(\mu^X)][var(Y)+(-\mu^Y - a)^2]+\\ \nonumber
  & \Phi(\mu^X)[var(Y)+(-\mu^Y + a)^2]\\  \nonumber
  =& [1-\Phi(\mu^X)][1+(\mu^Y + a)^2]+\Phi(\mu^X)[1+(\mu^Y -a )^2]   \\  \nonumber
  =& 1+(\mu^Y + a)^2 -4a\Phi(\mu^X)\mu^Y  \nonumber
\end{align}
where $a$ is given by $(1)$.

When $k=1$, the result is straight forward since the estimated centroid will approximately equal to
$0$, with error of size $\OhP(n^{-1/2})$. The cross-validation error
will be
\[
  \CV(1) = \frac{1}{m} \sum_{i=n+1}^{n+m} \| Y_i - \bar Y_n \|^2
         = 1 + (\mu^Y)^2 + \OhP(m^{-1/2}) + \OhP(n^{-1/2}).
\]
So if we have $1 + 2\Phi(\mu^Y)+ \frac{2\varphi(\mu^Y)}{\mu^Y} < 4\Phi(\mu^X)$, we have $\CV(2) < \CV(1)$
\end{proof}

We confirm this result with a simulation.  We perform $10$ replicates for each pair of $(\mu^X, \mu^Y)$, where both $\mu^X$ and $\mu^Y$ take value on grid of $[0,3]$ with step $0.1$.  In each
replicate, we generate $20000$ observations from two multivariate normal distributions with identity covariance, where one has mean $(\muX, \muY)$ and the other one has mean $(-\muX, -\muY)$.
We perform a single $2 \times 2$ fold of Gabriel cross-validation and report the
times (out of $10$ replicates) when $k=2$ is selected by the algorithm in stead of $k=1$. Figure~\ref{fig:overlap-color_plot} shows the frequency $k=2$ is selected by the algorithm for each pair of $(\mu^X, \mu^Y)$. The dark spot means high number (close to $10$) is selected by the algorithm, which means algorithm very likely will pick $k=2$ over $k=1$ for the corresponding $(\mu^X, \mu^Y)$. While light spot means algorithm prefer $k=1$ for the corresponding value of $(\mu^X, \mu^Y)$. We can see the simulation result perfectly align with the theoretical curve (the black line), which separates the $k=2$ zone from the $k=1$ zone. It demonstrates that the Gabriel cross-validation works exactly as it suppose to under such setting. The position of dark spots shows that when the two clusters are reasonably apart (not overlapping too heavily) in both dimensions, the Gabriel cross-validation is asymptotically consistent.

\begin{figure}[H]
\centering
\includegraphics[width=3in]{demo/overlap/color_plot.pdf}
\caption{Number of times $k=2$ is selected out of $10$ replicates for each pair of $(\muX, \muY)$. The heat map shows the frequency $k=2$ is selected by the algorithm, with light means low number (of $k=2$) is selected and dark color means high number is selected. The black line is the theoretical curve, above which the algorithm suppose to pick $k=2$ and below which algorithm select $k=1$.  }
\label{fig:overlap-color_plot}
\end{figure}

@@

>>>>

\newcommand{\Oh}{O}
\newcommand{\OhP}{O_p}

@@

\begin{proposition}\label{prop:self-consistency}

Let $\{ (X_i, Y_i) \}_{i=1}^{n+m}$ be the data from a single fold of Gabriel
cross-validation.  For any $k$, let $CV(k)$ be the cross-validation error for
this fold, computed as described in Section~\ref{sec:gabriel-cv-algorithm}.
We will assume that there are $K$ true centers $\mu(1), \dotsc,\mu(K)$, with
the $g$th cluster center partitioned as $\mu(g) = \bigl(\muX(g),
\muY(g)\bigr)$ for $g = 1, \dotsc, K$.  Suppose that
\begin{enumerate}[label=(\roman*)]
  \item \label{asn:self-consistency-noiseless}
    Each observation $i$ has a true cluster $G_i \in \{ 1, \dotsc, K \}$.
    There is no noise, so that $X_i = \muX({G_i})$ and $Y_i = \muY(G_i)$ for
    $i = 1, \dotsc, n+m$.
  \item \label{asn:self-consistency-distinct-mux}
    The vectors $\muX(1), \dotsc,\muX(K)$ are all distinct.
  \item \label{asn:self-consistency-distinct-muy}
    The vectors $\muY(1), \dotsc,\muY(K)$ are all distinct.
  \item \label{asn:self-consistency-train}
    The training set contains at least one member of each cluster: for all $g$
    in the range $1, \dotsc, K$, there exists at least one $i$ in the range
    $1, \dotsc, n$ such that $G_i = g$.
  \item \label{asn:self-consistency-test}
    The test set contains at least one member of each cluster: for all $g$ in
    the range $1, \dotsc, K$, there exists at least one $i$ in the range $n+1,
    \dotsc, n+m$ such that $G_i = g$.
\end{enumerate}
Then $CV(k) < CV(K)$ for $k < K$, and $CV(k) = CV(K)$ for $k > K$, so that
Gabriel CV correctly chooses $k = K$.
\end{proposition}

The result statement of Proposition~\ref{prop:self-consistency} follows from 
Lemmas~\ref{lem:self-consistency1} and~\ref{lem:self-consistency2}, which we
will now state and prove.

\begin{lemma}\label{lem:self-consistency1}
Suppose that the assumptions of Proposition~\ref{prop:self-consistency} are in
force.  If $k < K$, then $\CV(k) > 0$.
\end{lemma}
\begin{proof}
By definition,
\[
  \CV(k)
    =
      \sum_{i=n+1}^{n+m}
        \| Y_i - \bmuY (\hGX_i) \|^2,
\]
where $\bmuY(g)$ is the center of cluster $g$ returned from applying $k$-means
to $Y_1, \dotsc, Y_n$.  Assumptions~\ref{asn:self-consistency-noiseless}
and~\ref{asn:self-consistency-test}, imply that as $i$ ranges over the test
set $n+1, \dotsc, n+m$, the response $Y_i$ ranges over all distinct values in
$\{ \muY(1), \dotsc, \muY(K) \}$.
Assumption~\ref{asn:self-consistency-distinct-muy} implies that there are
exactly $K$ such distinct values.  However, there are only $k$ distinct values
of $\bmuY(g)$.  Thus, at least one summand
\(
  \| Y_i - \bmuY(\hGX_i) \|^2
\)
is nonzero.  Therefore,
\(
  \CV(k) > 0.
\)
\end{proof}

\begin{lemma}\label{lem:self-consistency2}
Suppose that the assumptions of Proposition~\ref{prop:self-consistency} are in
force.  If $k \geq K$, then $\CV(k) = 0$.
\end{lemma}
\begin{proof}
From assumptions~\ref{asn:self-consistency-noiseless},
\ref{asn:self-consistency-distinct-muy},
and~\ref{asn:self-consistency-train}, we know the cluster centers
gotten from applying $k$-means to $Y_1, \dotsc, Y_n$ must include
$\muY(1), \dotsc, \muY(K)$.  Without loss of generality, suppose that
$\bmuY(g) = \muY(g)$ for $g = 1, \dotsc, K$.  This implies that
$\hGY_i = G_i$ for $i = 1, \dotsc, n$.  Thus, employing
assumption~\ref{asn:self-consistency-noiseless} again, we get that
$\bmuX(g) = \muX(g)$ for $g = 1, \dotsc, K$.

@@

Proposition~\ref{prop:self-consistency} tells us that the Gabriel cross-validation
method recovers the true number of clusters when the noise is negligible.
While this result gives us some assurance that the procedure is well-behaved,
we can bolster our confidence and gain insight into its workings by analyzing
its behavior in the presence of noise.  We first study the case of a single
cluster in two dimensions with correlated Gaussian noise.

\begin{proposition}\label{prop:single-2d}
Suppose that $\{ (X_i, Y_i) \}_{i=1}^{n + m}$ is data from a single fold
of Gabriel cross-validation, where each $(X,Y)$ pair in $\R^2$ is an
independent draw from a mean-zero multivariate normal distribution with unit
marginal variances and correlation $\rho$.  In this case, the data are drawn
from a single cluster; the true number of clusters is~$1$.  If $|\rho| < 1/2$,
and $k > 1$, then $\CV(1) < \CV(k)$ with probability tending to one as $m$
and $n$ increase.
\end{proposition}

@@

Intuitively, when the correlation is high, the response feature, $Y$, looks
similar to the predictor feature, $X$. Prediction error on $X$ always
decreases with larger $k$. Thus, when the correlation is high, the prediction
error for $Y$ will also decrease with larger $k$. This explains why
cross-validation breaks down in the presence of strong correlation.

In Appendix~\ref{sec:single-general-dim}, using similar techniques to those used
to prove Proposition~\ref{prop:single-2d}, we derive an analogous result for correlated
Gaussian noise in more than two dimensions. A similar phenomenon holds:
the Gabriel cross-validation method fails when the first principal component
of the $Y$ variables is strongly correlated with a linear combination of the
$X$ variables.

@@

We will now analyze a simple two-cluster setting, and derive conditions for
Gabriel cross-validation to correctly prefer $k=2$ clusters to $k=1$. In the
statement of the result, $\Phi(\cdot)$ and $\varphi(\cdot)$ denote the
standard normal cumulative distribution function and density, respectively.

\begin{proposition}\label{prop:twoclust}
Suppose that $\{(X_i,Y_i)\}_{i=1}^{n+m}$ is data from a single fold of Gabriel
cross-validation, where each $(X,Y)$ pair in $\R^2$ is an independent draw
from an equiprobable mixture of two multivariate normal distributions with
identity covariance. Suppose that the first mixture component has mean
$\mu = (\muX, \muY)$ and the second has mean $-\mu = (-\muX, -\muY)$,
where $\muX \geq 0$ and $\muY \geq 0$. 
If 
\(
  2 \varphi(\muY) + \muY + 2 \muY \Phi(\muY) < 4 \muY \Phi(\muX),
\)
then $\CV(2) < \CV(1)$ with probability tending to one as $m$ and $n$ increase.
\end{proposition}

\begin{proof}
There are two clusters: observations from cluster~$1$ are distributed as
\(
  \mathcal{N}(\mu, I)
\)
and observations from cluster~$2$ are distributed as
\(
\mathcal{N}(-\mu, I)
\)
where $\mu = (\muX, \muY)$.  Without loss of generality, $\muX \geq 0$ and
$\muY \geq 0$.  Let $G_i$ be the true cluster of observation~$i$
where, by assumption,
\[
  \Pr(G_i=1) = \Pr(G_i=2) = 1/2.
\]
After applying $k$-means to $\{ Y_i\}_{i=1}^{n}$ with $k=2$, if $n$ is large
enough, then the estimated cluster means $\bar{\mu}^Y_1$ and $\bar{\mu}^Y_2$
will be close to $\E(Y \mid Y>0)$ and $\E(Y \mid Y < 0)$, with errors of
size $\OhP(n^{-1/2})$. To compute these quantities, let
$(X_1, Y_1) \sim \mathcal{N}(\mu, I)$ and $(X_2, Y_2) \sim \mathcal{N}(-\mu, I)$
be draws from the mixture components, and let $(X,Y)$ be defined such that
$\Pr(X= X_1, Y = Y_1) = \Pr(X= X_2, Y = Y_2) = 1/2$. Then,
\begin{align*}
  \E(Y \mid Y>0)
  &= \E(Y_1 \mid Y_1 > 0) \cdot \Pr(Y = Y_1 \mid Y >0)
   + \E(Y_2 \mid Y_2 > 0) \cdot \Pr(Y = Y_2 \mid Y > 0)
\\
&=  \{ \muY + \varphi(\muY) / \Phi(\muY) \} \cdot \Phi(\muY)
+ [-\muY + \varphi(\muY) / \{ 1 - \Phi(\muY) \}] \cdot \{ 1 - \Phi(\muY) \}
\\
  &= 2\varphi(\muY)+ 2\muY\Phi(\muY)- \muY. 
\end{align*}
In the second line, we have used
Lemma~\ref{lem:truncated-normal-moments} from
Appendix~\ref{app:technical-lemmas} to compute the conditional expectations;
$\varphi()$ and $\Phi()$ are the standard normal density and
cumulative distribution function, respectively. By symmetry,
\begin{align*}
  \E(Y \mid Y<0) &= -\E(Y \mid Y > 0).
\end{align*}

The classification rule learned from the training data
$\{(X_i, \hGY_i)\}_{i=1}^{n}$ will have its decision boundary at $0 + \OhP(n^{-1/2})$;
that is, in the limit,
observations will get classified as coming from cluster~1 when $X > 0$. 
Set $a = \E(Y \mid Y > 0)$.
Up to terms of order $\OhP(n^{-1/2})$,
the cross-validation error from a single observation is distributed as
\[
  (Y - a)^2 1\{ X > 0 \} + (Y + a)^2 1\{X < 0\}.
\]
Using the fact that conditional on the mixture component, the $X$ and $Y$
coordinates are independent, we can compute the expectation of the first
summand as
\begin{align*}
  \E[(Y - a)^2 1\{ X > 0 \}]
  &= (1/2) \E[(Y_1 - a)^2] \Pr(X_1 > 0)
  + (1/2) \E[(Y_2 - a)^2] \Pr(X_2 > 0)
\\
&= (1/2) \Phi(\muX) \E(Y_1 - a)^2  + (1/2) \{ 1 - \Phi(\muX) \} \E(Y_2 - a)^2
\\
&=
(1/2) [ 1  + \Phi(\muX) (\muY - a)^2 + \{ 1 - \Phi(\muX) \} (-\muY - a)^2 ].
\end{align*}
By a similar calculation, the expectation of the second summand is
\[
  \E[(Y + a)^2 1\{ X < 0 \}]
  =
  (1/2)[ 1 + \{ 1 - \Phi(\muX) \} (\muY + a)^2 + \Phi(\muX) (-\muY + a)^2 ].
\]
Adding the two terms, we get that the expected cross-validation error from a
single observation is
\[
  1 + \Phi(\muX) (\muY - a)^2 + \{ 1 - \Phi(\muX) \} (\muY + a)^2
    =
  1 + (\muY)^2 + a \, \{  a + 2 \muY - 4 \muY \Phi(\muX) \}.
\]
Thus, the $k = 2$ cross-validation error on the test set is
\begin{align*}
  \CV(2)
  &=
    \frac{1}{m}
    \sum_{i=n+1}^{n+m}
      \| (Y_i - \bmuY_1) 1\{\hGX_i = 1\} \|^2
      +
      \| (Y_i - \bmuY_2) 1\{\hGX_i = 2\} \|^2
\\
&=
1 + (\muY)^2 +  a \, \{ a + 2 \muY - 4 \muY \Phi(\muX) \}
+ \OhP(n^{-1/2}) + \OhP(m^{-1/2}).
\end{align*}

When $k=1$, the $k$-means centroid is equal to the sample mean $\bar Y_n = (1/n)
\sum_{i=1}^{n} Y_i$, approximately equal to 
$\E(Y) = 0$, with error of size $\OhP(n^{-1/2})$. The cross-validation error is
\[
  \CV(1) = \frac{1}{m} \sum_{i=n+1}^{n+m} \| Y_i - \bar Y_n \|^2
         = 1 + (\mu^Y)^2 + \OhP(m^{-1/2}) + \OhP(n^{-1/2}).
\]
Thus, if $a + 2 \muY - 4 \muY \Phi(\muX) < 0$, then $\CV(2) < \CV(1)$ with
probability tending to one as $m$ and $n$ increase. Substituting the
expression for $\E(Y \mid Y > 0)$ in place of $a$, the inequality holds
precisely when
\(
  2 \varphi(\muY) + \muY + 2 \muY \Phi(\muY) < 4 \muY \Phi(\muX).
\)
\end{proof}

We confirm the result of Proposition~\ref{prop:twoclust} with a simulation.
We perform $10$ replicates for each $(\mu^X, \mu^Y)$ pair, sweeping over
a two-dimensional grid of values in the domain $[0,3] \times [0,3]$, with step
size $0.1$ in each dimension.  In each
replicate, we generate $N=20000$ observations from an equiprobably mixture
of multivariate normal distributions with identity covariance, with one
component having mean $(\muX, \muY)$ and
the other component having mean $(-\muX, -\muY)$. We perform a single $2 \times 2$
fold of Gabriel cross-validation and report the number of times
(out of $10$ replicates) where $k=2$ is selected by the algorithm instead of $k=1$.
Figure~\ref{fig:overlap-color_plot} shows the frequency with which $k=2$ is selected by
the algorithm for each $(\mu^X, \mu^Y)$ pair. Darker (red) colors indicate
higher numbers (close to $10$), situations where $k = 2$ is selected more
often than $k = 1$. Ligher (blue) colors indicate that $k = 1$ is preferred.
We can see the simulation result perfectly align with the
theoretical curve (the black line), which separates the $k=2$ zone from the
$k=1$ zone.

\begin{figure}
\centering
\includegraphics[width=3in]{demo/overlap/color_plot.pdf}
\caption{Number of times $k=2$ is selected out of $10$ replicates for each
pair of $(\muX, \muY)$. The heat map shows the frequency $k=2$ is selected by
the algorithm, with light (blue) means $k = 1$ is preferred to $k = 2$, and
dark (red) indicates $k = 2$ is preferred to $k = 1$.
The black line is the theoretical boundary determined from
Proposition~\ref{prop:twoclust}.}
\label{fig:overlap-color_plot}
\end{figure}

@@

\section{Analysis of Gabriel method: Single cluster in more than two dimensions}
\label{sec:single-general-dim}

Let $\Sigma_{YY} = U \Lambda U^T$  be the eigendecomposition of $\Sigma_{YY}$,
with leading eigenvalue $\lambda_1$ and corresponding eigenvector $u_1$. Then
the centroid of $k$-means applying on $(y_1,..,y_n)$ is on the first principal
component
of $Y$,\[	E(u^T_1 Y|u^T_1 Y>0) = \bar{\mu}^Y_1 =\sqrt{2 \lambda_1/\pi}u_1\] and 
\[	E(u^T_1 Y|u^T_1 Y<0) = \bar{\mu}^Y_2 =-\sqrt{2 \lambda_1/\pi}u_1\]
 where $u^T_1 Y \sim \mathcal{N}(0,\lambda_1)$.
